Higher-Order unification for free!

### 2024-04-12 paper/main.tex
a pass on 7 and 9

% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}. 
There exists systems
such as Teyjus~\cite{Nadathur2001} and $\lambda$Prolog~\cite{lamProlog} which 
delay the resolution of \notllambda unification problems if the substitution
is not able to put them in \llambda.

In this section we want to show how we can adapt the unification of the object
language in the meta language by simply adding (or removing) rules to the
\elpiIn{progress} predicate.
%
% make test ONLY=7005 TEX=tex
\printAlll
  {{{(X\appsep a),a},
    {X,\lambda x.a}}}
  {{}}
  {{}}
  {{}}

In the example above, $\foUnifPb_1$ is in \notllambda. \emph{If} the object
language delays the first unification problem waiting $X$ to be
be instantiated in a future unification, we can relax \cref{def:progBetaFail}. 
Instead of failing because the \lhs of the considered \linkbeta $l$ becomes rigid,
we keep it in \linkStore until the head of its \rhs also become rigid.
In this case, since \lhs and \rhs have rigid heads, they can be unified
just before removing $l$ from \linkStore. We can note that this rule trivially
guarantees \fullref{prop:fidelity}. On the other hand, the occur check becomes 
partial: there exists \linkbeta with a non-flexible \lhs.

A second strategy to deal with problem that are in \notllambda is to make
approximations. This is the case for example of the unification algorithm of Coq
used in its type-class solver~\cite{sozeau08}. The approximation consists in
forcing a choice (among the others) when the unification problem is outside
\llambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
that this is of course an approximation, since $\sigma = \{X \mapsto \lambda x.Y, Y \mapsto
\_\}$ is another valid substitution for the original problem. We stress the 
fact that, again, our unification procedure in the meta language can be 
accommodated for this new behavior: given a \linkbeta, if \lhs is not in \llambda,
then \elpiIn{progress} can try to align the rightmost arguments and unify
the resulting heads.

Note that delaying unification outside \llambda can leave \linkbeta during the
decompilation phase. Therefore, new rules to \elpiIn{commit-links} should
be added accordingly.

% \noteD{Paragraph for decompile in case of \maybebeta}
% \todo{cita paper su estensioni a llam via sospensione etc}
\todo[inline]{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}

@@

Different strategies can be used to unify terms of the object language in the
meta langauge. One initial approach involves integrating \Uo as a procedure
within the ML, outlined as follows:

@@

Opting for this method would result in a suboptimal utilization of the logic
programming engine provided by the meta language, as it degrades indexing by
eliminating data from rule heads. Additionally, embedding \Uo within the meta
language is likely to be significantly slower compared to a built-in
solution.

% Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
Another possibility is to avoid having the application and abstraction nodes
in the syntax tree, and use the ones of meta language, as in the following:

@@

\noindent
However, two reasons justify rejecting this encoding. Firstly, in CIC, it is
not always feasible to adopt it due to the meta language's type system being
too limited to accommodate that of the object language. \\
Secondly, the CIC encoding provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples.

In the literature we could find related encoding of the Calculus of
Constructions~\cite{felty93lics}. The goal of that work was to exhibit
a logic program performing proof checking in CC and hence relate the
proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
programs) with the Calculus of Constructions. The encoding is hence tailored
toward a different goal, and utilizes three relations to represent the
equational theory of CC. Section 6 contains a discussion about the use of the
unification procedure of the meta language in presence of non ground goals, but
the authors do not aim at exploiting it to the degree we want.\todo{sucks}

@@

Our encoding provides a third option that addresses all the considerations
mentioned earlier. It capitalizes on the benefit of not requiring the recoding
of the unification algorithm of the object language. Instead, it employs the
unification capabilities of the meta language, facilitated by the various links
we establish to manage ``problematic'' subterms. Moreover, our encoding takes
advantage of indexing algorithms and mode analysis for static clause filtering. It's noteworthy
that we only replace the minimum necessary information with variables,
specifically targeting \maybeeta and \notllambda subterms.

Our proposed approach is highly adaptable to
align closely with the behavior of the object language. It is not tightly
coupled with the Coq system but can serve as a flexible framework for meta
programming in any ML.

Furthermore, the unification process we propose is tailored for potential future
implementations of tabled search, incorporating memoization to retrieve
solutions from previous searches.

>>>>

% from https://www.cs.mcgill.ca/~bpientka/papers/unif_miller60.pdf
Working with terms in \llambda is sometime too restrictive \cite{Abel2018ExtensionsTM}
and a there exist a few strategies to go bejond \llambda without 
implementing Huet's algorithm~\ref{REF}.

Some implementations
of $\lambda$Prolog~\cite{lamProlog} 
such as Teyjus~\cite{Nadathur2001} delay the resolution of \notllambda unification problems until the substitution
 put them in \llambda. Other systems,
for example of the unification algorithm of Coq
used in its type-class solver~\cite{sozeau08},
 apply heuristics like preferring projection over mimic, and commit
 to that solution.

In this section we show how we can implement these strategies
by simply adding (or removing) rules to the
\elpiIn{progress} predicate.
In the example below $\foUnifPb_1$ is in \notllambda.
%
% make test ONLY=7005 TEX=tex
\printAlll
  {{{(X\appsep a),a},
    {X,\lambda x.a}}}
  {{}}
  {{}}
  {{}}

If we want the object language unification to delay the first unification problem
(waiting for $X$ to be be instantiated by a future unification),
we can relax \cref{def:progBetaFail}. 
Instead of failing because the \lhs of the considered \linkbeta $l$ becomes rigid,
we keep it in \linkStore until the head of its \rhs also become rigid.
In this case, since \lhs and \rhs have rigid heads, they can be unified
just before removing $l$ from \linkStore. We can note that this rule trivially
guarantees \fullref{prop:fidelity}. On the other hand, the occur check becomes 
partial since we break \fullref{inv:linklhs}. %: there exists \linkbeta with a non-flexible \lhs.

If we want to follow the second strategy and pick an
arbitrary solution. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
argument of the two terms is the same and unification can succeed by assigning
$X a$ to $Y$. Note
that this is not the only solution, since $\sigma = \{X \mapsto \lambda x.Y \}$
is another valid substitution. Our unification procedure
can be modified to accommodate this behavior by adding a
rule to \elpiIn{progress} that tries to
align the rightmost arguments and unify
the resulting heads whenever \lhs of a \linkbeta is not in \llambda.

Note that delaying unification outside \llambda can leave \linkbeta for the
decompilation phase. Therefore \elpiIn{commit-links} should
be modified accordingly.

% \noteD{Paragraph for decompile in case of \maybebeta}
% \todo{cita paper su estensioni a llam via sospensione etc}
\todo{cita teyjus (1) era 2nd order HO (huet's algorithm), teyjus 2 è llam ma sospende i disagreement pairs fuori da llam}

@@

Different strategies can be used to unify terms of the object language.
% in the meta langauge. 
The first approach that comes to mind consist in implementing \Uo as a
regular routine in the ML as follows:

@@

\noindent
Opting for this method would result in a suboptimal utilization of the logic
programming engine provided by the meta language, as it degrades indexing by
eliminating all data from rule heads. Additionally,
implementing an unification procedure in the meta
language is likely to be significantly slower compared to the built-in one.

% Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.
Another possibility is to avoid having application and abstraction nodes
in the syntax tree, and use the ones of meta language, as in the following:

@@

\noindent
However, this encoding has two big limitations. Firstly, in CIC, it is
not always feasible to adopt it due to the meta language's type system being
too limited to accommodate for that one the object language. E.g. CIC
can typecheck variadic functions, see for example~\ref{CPDT chlipala}. \\
Secondly, the CIC encoding provided by Elpi is primarily utilized for meta
programming, in order to extend the Coq system. Consequently, it must be able to
manipulate terms that are not known in advance
without relying on introspection primitives such as Prolog's
\texttt{functor} and \texttt{arg}. In this context, constants need to live in
an open world, akin to the \elpiIn{string} data type used in the preceding
examples.

In the literature we could find a related encoding of the Calculus of
Constructions (CC)~\cite{felty93lics}. The goal of that work was to exhibit
a logic program performing proof checking for CC and hence relate the
proof system of intuitionistic higher-order logic (that animates $\lambda$Prolog
programs) with the one of CC. The encoding is hence tailored
toward a different goal, for example it utilizes three relations to represent the
equational theory of CC. Section 6 contains a discussion about the use of the
unification procedure of the meta language in presence of non ground goals, but
the authors do not aim at exploiting it to the degree we want.\todo{sucks}

@@

\emph{The approach presented in this paper}
provides a third option that addresses all the concerns
mentioned earlier. It capitalizes on the benefit of not requiring to fully
implement
the unification algorithm of the object language. Instead, it employs the
unification capabilities of the meta language, facilitated by the various links
%we establish
to manage ``problematic'' subterms. As a result of this choice our encoding takes
advantage of indexing data structures and mode analysis for clause filtering.
It it worth mentioning that we only replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable by the meta language logic programming engine.
Moreover, the unification process we obtain is ready to take
advantage of potential improvements to the programming engine,
such as tabled search, and apply forms of static analysis for
the meta language, such as determinacy, to the object language. 
Finally, our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment, and
it is not tightly coupled with CIC.
